---
title: "Comparing DNA Reads with Zooplankton Counts"
author: "Analysis Notebook"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Introduction

This notebook compares DNA sequencing reads (eDNA) with traditional zooplankton microscopy counts. The goal is to understand how well DNA-based methods match traditional counting methods for different taxonomic groups.

# Load Required Packages

```{r load-packages}
# Install packages if needed (uncomment the lines below)
# install.packages(c("tidyverse", "ggplot2", "scales"))

# Load packages
library(tidyverse)  # For data manipulation and plotting
library(scales)     # For nice axis formatting
```

# Load and Prepare Data

```{r load-data}
# Read the CSV file
data <- read.csv("asvtaxonkelp.csv", stringsAsFactors = FALSE)

# Look at the structure
cat("Total rows in dataset:", nrow(data), "\n")
cat("Column names:", paste(names(data), collapse = ", "), "\n\n")

# Show first few rows
head(data) %>% 
  select(Sample_code, phylum, zooplank_abundance, eDNA_reads, total_reads) %>%
  print()
```

## Filter for Zooplankton Data

```{r filter-zooplankton}
# Keep only zooplankton records (where we have count data)
zoo_data <- data %>%
  filter(!is.na(zooplank_abundance) & zooplank_abundance >= 0,
         !is.na(eDNA_reads),
         !is.na(total_reads),
         total_reads > 0) %>%
  # Focus on key zooplankton phyla
  filter(phylum %in% c("Arthropoda", "Annelida", "Mollusca", "Cnidaria", 
                       "Bryozoa", "Echinodermata", "Chordata", "Rotifera"))

cat("Filtered to", nrow(zoo_data), "zooplankton records\n")
cat("Unique phyla:", paste(unique(zoo_data$phylum), collapse = ", "), "\n")
```

## Calculate Relative Abundances

```{r calculate-relative}
# Calculate relative abundances
zoo_data <- zoo_data %>%
  mutate(
    # Relative DNA reads (already calculated in dataset, but we recalculate for clarity)
    rel_DNA = eDNA_reads / total_reads,
    
    # Relative zooplankton abundance (proportion within each sample)
    rel_zoo = zooplank_abundance / Total_Zooplankton,
    
    # Log10 transform (adding small value to avoid log(0))
    log_rel_DNA = log10(rel_DNA + 1e-10),
    log_rel_zoo = log10(rel_zoo + 1e-10)
  )

# Summary statistics
cat("\nSummary of relative abundances:\n")
summary(zoo_data[, c("rel_DNA", "rel_zoo")])
```

## Aggregate by Taxonomic Group

For clearer visualization, we'll aggregate data by phylum for each sample.

```{r aggregate-by-phylum}
# Sum abundances by sample and phylum
phylum_data <- zoo_data %>%
  group_by(Sample_code, phylum, Total_Zooplankton, total_reads) %>%
  summarize(
    zoo_count = sum(zooplank_abundance, na.rm = TRUE),
    DNA_reads = sum(eDNA_reads, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Calculate relative abundances
    rel_zoo = zoo_count / Total_Zooplankton,
    rel_DNA = DNA_reads / total_reads,
    
    # Log10 transform
    log_rel_zoo = log10(rel_zoo + 1e-10),
    log_rel_DNA = log10(rel_DNA + 1e-10)
  )

cat("Aggregated to", nrow(phylum_data), "phylum-level records\n")
cat("Samples:", length(unique(phylum_data$Sample_code)), "\n")
cat("Phyla:", length(unique(phylum_data$phylum)), "\n")
```

# Analysis 1: Per-Group Alignment Strength

This analysis calculates the correlation between DNA reads and zooplankton counts for each taxonomic group.

```{r alignment-strength}
# Calculate Spearman correlation for each phylum
alignment_stats <- phylum_data %>%
  group_by(phylum) %>%
  summarize(
    n_samples = n(),
    spearman_rho = cor(log_rel_zoo, log_rel_DNA, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(spearman_rho))

print(alignment_stats)
```

## Plot: Alignment Strength by Phylum

```{r plot-alignment, fig.height=6}
# Reorder phyla by correlation strength
alignment_stats <- alignment_stats %>%
  mutate(phylum = fct_reorder(phylum, spearman_rho))

# Create the plot
ggplot(alignment_stats, aes(x = spearman_rho, y = phylum)) +
  geom_point(size = 4) +
  labs(
    title = "Per-group alignment strength",
    x = "Spearman rho (log10 rel zoo vs log10 rel reads)",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 12)
  )
```

**Interpretation:** This plot shows how well DNA reads correlate with microscopy counts for each group. Higher values (closer to 1) mean better agreement between methods.

# Analysis 2: Overall Relationship

Now let's look at the overall relationship between DNA reads and zooplankton counts across all samples and taxa.

```{r overall-correlation}
# Calculate overall correlation
overall_cor <- cor(phylum_data$log_rel_zoo, phylum_data$log_rel_DNA, 
                   method = "spearman", use = "complete.obs")

cat("Overall Spearman correlation:", round(overall_cor, 3), "\n")
```

## Plot: DNA Reads vs Zooplankton Abundance

```{r plot-overall, fig.height=7, fig.width=10}
# Create color palette for phyla
phylum_colors <- scales::hue_pal()(length(unique(phylum_data$phylum)))
names(phylum_colors) <- sort(unique(phylum_data$phylum))

# Plot with linear fit
ggplot(phylum_data, aes(x = log_rel_zoo, y = log_rel_DNA, color = phylum)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  scale_color_manual(values = phylum_colors, name = "Taxa") +
  labs(
    x = "Log10 relative zooplankton abundance",
    y = "Log10 relative reads"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

**Interpretation:** This scatterplot shows the overall relationship. Each point is one phylum in one sample. The black line shows the trend - a positive slope means DNA reads tend to increase with microscopy counts.

# Analysis 3: Relationship by Taxonomic Group

Let's examine the relationship separately for each phylum to see if some groups show better agreement than others.

```{r plot-by-phylum, fig.height=10, fig.width=12}
# Plot with separate trend lines for each phylum
ggplot(phylum_data, aes(x = log_rel_zoo, y = log_rel_DNA, color = phylum)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = phylum_colors, name = "Taxa") +
  labs(
    x = "Log10 relative zooplankton abundance",
    y = "Log10 relative reads"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

**Interpretation:** This plot shows separate trend lines for each taxonomic group, revealing which groups show positive, negative, or no relationship between DNA and microscopy data.

# Analysis 4: Faceted View by Taxonomic Group

For a clearer view of individual groups, let's create separate panels for each phylum.

```{r plot-faceted, fig.height=10, fig.width=12}
# Calculate stats for each phylum to display on plots
phylum_stats <- phylum_data %>%
  group_by(phylum) %>%
  summarize(
    rho = cor(log_rel_zoo, log_rel_DNA, method = "spearman", use = "complete.obs"),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("ρ = ", round(rho, 2), "\nn = ", n)
  )

# Join stats to data
phylum_data_with_stats <- phylum_data %>%
  left_join(phylum_stats, by = "phylum")

# Create faceted plot
ggplot(phylum_data_with_stats, aes(x = log_rel_zoo, y = log_rel_DNA)) +
  geom_point(aes(color = phylum), size = 2, alpha = 0.7, show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1) +
  geom_text(aes(label = label), x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, size = 3.5) +
  facet_wrap(~ phylum, scales = "free", ncol = 3) +
  labs(
    x = "Log10 relative zooplankton abundance",
    y = "Log10 relative reads"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold")
  )
```

**Interpretation:** Each panel shows one taxonomic group. The blue line is the best-fit trend. The correlation coefficient (ρ) and sample size (n) are shown in each panel.

# Summary Statistics

```{r summary-table}
# Create summary table
summary_table <- phylum_data %>%
  group_by(phylum) %>%
  summarize(
    n_samples = n(),
    mean_zoo_abundance = mean(zoo_count, na.rm = TRUE),
    mean_DNA_reads = mean(DNA_reads, na.rm = TRUE),
    correlation = cor(log_rel_zoo, log_rel_DNA, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  ) %>%
  arrange(desc(correlation))

print(summary_table)
```

# Key Findings

```{r findings}
cat("\n=== KEY FINDINGS ===\n\n")

cat("1. Overall correlation between DNA reads and zooplankton counts:\n")
cat("   Spearman ρ =", round(overall_cor, 3), "\n\n")

cat("2. Phyla with strongest agreement (top 3):\n")
top_3 <- alignment_stats %>% arrange(desc(spearman_rho)) %>% head(3)
for(i in 1:3) {
  cat("   ", i, ". ", top_3$phylum[i], " (ρ = ", round(top_3$spearman_rho[i], 2), 
      ", n = ", top_3$n_samples[i], ")\n", sep = "")
}

cat("\n3. Phyla with weakest agreement (bottom 3):\n")
bottom_3 <- alignment_stats %>% arrange(spearman_rho) %>% head(3)
for(i in 1:3) {
  cat("   ", i, ". ", bottom_3$phylum[i], " (ρ = ", round(bottom_3$spearman_rho[i], 2), 
      ", n = ", bottom_3$n_samples[i], ")\n", sep = "")
}

cat("\n4. Sample size: ", length(unique(phylum_data$Sample_code)), " samples\n", sep = "")
cat("5. Number of phyla analyzed: ", length(unique(phylum_data$phylum)), "\n\n", sep = "")
```

# Conclusion

This analysis compares eDNA sequencing with traditional microscopy counts for zooplankton. The results show varying levels of agreement across different taxonomic groups, which helps us understand:

- **Which groups** are well-detected by DNA methods
- **Which groups** may require method refinement
- **Overall patterns** in the relationship between DNA and morphological identification

The correlation values (Spearman's rho) range from negative to positive, indicating that:
- **Positive correlations**: DNA reads increase with microscopy counts (good agreement)
- **Near-zero correlations**: Little relationship between methods
- **Negative correlations**: Inverse relationship (requires investigation)

---

**How to use this notebook:**

1. Make sure the CSV file `asvtaxonkelp.csv` is in the same directory
2. Install required packages if needed (see first code chunk)
3. Run all chunks sequentially (Ctrl/Cmd + Shift + Enter)
4. Adjust the phyla filter if you want to focus on different groups
5. Modify plot aesthetics as needed for publication

---
