---
title: "R Notebook"
output: html_notebook
---
```{r}
```{r}
# 0) Libraries & global theme (no grid lines)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(scales)
library(patchwork)
library(janitor)
library(forcats)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid       = element_blank(),
      axis.title       = element_text(size = 14),
      axis.text        = element_text(size = 14),
      legend.title     = element_text(size = 14),
      legend.text      = element_text(size = 14),
      strip.text       = element_text(size = 14)
    )
)

# 1) Read & compute per‐cm² abundances

amphi <- read.csv("amphi_heatmap.csv") %>%
  clean_names() %>%
  filter(site != "Scalpay") %>% 
  mutate(amphipod_p_cm = amphipod / for_standardizing,
         caprellidae_p_cm = caprellidae / for_standardizing,
         gammaridae_p_cm = gammaridae / for_standardizing,
         jassa_p_cm = jassa / for_standardizing,
         corophium_p_cm = corophium / for_standardizing)

```